### 2019-12-10 boek.tex
updates

\PassOptionsToPackage{table}{xcolor}
\documentclass[8pt]{book}

@@

\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm]{geometry}

\usepackage[
	% set width and height to a4 width and height + 6mm
	%width=16truecm, height=24truecm,
	a4,
	% use any combination of these options to add different cut markings
	cam, axes, frame, cross,
	% set the type of TeX renderer you use
	pdftex,
	% center the contents
	center
]{crop}

% adjust margins (should be done before importing fancyhdr!) %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm

@@

% sections in allcaps
\usepackage{titlesec}
%\titleformat{\section} {\Large\bfseries\center}{}{0em}{\hrule height 0.4pt\vspace{2pt}}[\hrule height 0.4pt] \titlespacing*{\section} {0pt}{0pt}{10pt}
%\titleformat{\subsection} {\large\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsection} {0pt}{0pt}{0pt}
%\titleformat{\subsubsection} {\normalsize\bfseries\center}{}{0em}{\vspace{0pt}} \titlespacing*{\subsubsection}  {0pt}{0pt}{0pt}
%\titleformat{\section} [block]{\Large\bfseries\filcenter}{}{0em}{\MakeUppercase}
%\titleformat{\subsection} [block]{\large\bfseries\filcenter}{}{0em}{\MakeUppercase}
%\titleformat{\subsubsection} [block]{\normalsize\bfseries\filcenter}{}{0em}{\MakeUppercase}

@@

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE HEADER %%%
%%%%%%%%%%%%%%%%%%%%%
\usepackage{fancyhdr}  
\pagestyle{fancy} 
\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
\fancypagestyle{frontmatter}{%    
	\fancyhf{} 	
	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
	\fancypagestyle{plain}{
		\renewcommand{\headrulewidth}{0pt}
		\fancyhead{}
		\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	} 
}
\fancypagestyle{mainmatter}{%    
	\fancyhf{} 	
	\fancyhead[RO,LE]{\nouppercase{\small \textsf{\leftmark}}}
	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
}
\setlength{\headheight}{15pt}

@@

\definecolor{light-gray}{gray}{0.90}
\definecolor{med-gray}{gray}{0.60}
\tikzset{
	thumb/.style={
		%   draw=black,
		fill=light-gray,
		text=black,
		minimum height=\thumbheight, %\thumbheight,
		text width=\thumbwidth,
		outer sep=0pt,%   outer sep=10pt,
		font=\sffamily\Huge,
	}
}
\newcommand{\oddthumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north east,] at ($%
	(current page.north east)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
\newcommand{\eventhumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north west,] at ($%
	(current page.north west)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
% create a new command to set a new lettergroup
\newcommand{\lettergroup}[1]{%
	\fancyhead[LO]{\oddthumb{#1}}%
	\fancyhead[RE]{\eventhumb{#1}}%
	\fancypagestyle{chapterstart}{%
		\renewcommand{\headrulewidth}{0pt}
		\renewcommand{\footrulewidth}{0pt}
		\fancyhf{}
		\chead{\oddthumb{#1}}% chapters start only on odd pages
		\fancyfoot[RO,LE]{\textsf{Page \thepage}} %\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	}
	\thispagestyle{chapterstart}
	\stepcounter{letternum}%
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE FIGURE HELPERS %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\hugefigure}{\textwidth}
\newcommand{\LARGEfigure}{0.8\textwidth} % 1 per col
\newcommand{\LARgefigure}{0.6\textwidth} % 1 per col
\newcommand{\Largefigure}{0.5\textwidth} % 1 per col
\newcommand{\largefigure}{0.37\textwidth} % 2 per col
\newcommand{\normalfigure}{0.3\textwidth} % 3 per col
\newcommand{\smallfigure}{0.24\textwidth} % 3 per col
\newcommand{\Smallfigure}{0.19\textwidth} % 4 per col
\newcommand{\SMALLfigure}{0.15\textwidth} % 5 per col

\makeatletter
\newcommand*{\centerfloat}{%
	\parindent \z@
	\leftskip \z@ \@plus 1fil \@minus \textwidth
	\rightskip\leftskip
	\parfillskip \z@skip}
\makeatother

\newcommand{\picfig}[4]{
\begin{figure}[ht!]
	\centering \centerfloat
	\includegraphics[width=#2]{#1}
	\caption{#4}
	\label{#3}
\end{figure}}

@@

\begin{document}

\pagenumbering{alph}

\frontmatter
\pagestyle{empty}

@@

%{\large \textbf{Abstract}}
%\begin{quote}
	%Recent developments in single-cell transcriptomics have opened new opportunities for studying dynamic processes in a high-throughput and unbiased manner. Starting from a mixture of cells in different stages of a developmental process, unsupervised trajectory inference algorithms aim to automatically reconstruct the underlying developmental path that cells are following. In this review, we break down the strategies used by this novel class of methods, and organize their components into a common framework, highlighting several practical advantages and disadvantages of the individual methods. We also give an overview of new insights these methods have already provided regarding the wiring and gene regulation of cell differentiation. As the trajectory inference field is still in its infancy, we propose several future developments which will ultimately lead to a global and data-driven way of studying immune cell differentiation.
%\end{quote}

\vfill
\newpage
\graphicspath{{./ch1_introduction/}}
\include{ch1_introduction/content}

@@

\vfill
{\large \textbf{Author contributions}}
\begin{quote}
	$\bullet$ \textbf{R.C.} and W.S. designed the study.\\
	$\bullet$ \textbf{R.C.}, W.S., and L.D. performed the experiments and analysed the data.\\
	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
	$\bullet$ \textbf{R.C.} wrote the original manuscript.\\
	$\bullet$ \textbf{R.C.}, W.S., L.D., and Y.S. reviewed and edited the manuscript. \\
	$\bullet$ Y.S. supervised the project.
\end{quote}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%   CH4: Self-assessment    %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage{\thispagestyle{empty}\cleardoublepage}
\chapter{Self-assessment in trajectory inference} 
\chaptermark{Self-assessment in trajectory inference.} % short title
\label{chap:selfassessment}

{\large \textbf{Abstract}}
\begin{quote}
	\textbf{Purpose:} A brief of 75 articles of trajectory tools reveals 
	that only 37\% of works contain a quantitative self-assessment. 
	While self-assessments are universally biased in favour of the
	authors, it is dangerous to publish a computational tool
	without quantitatively demonstrating its performance compared to
	state-of-the-art methods. \\
	\textbf{Results:} We hypothesise that low self-assessment rates are
	primarily caused by a lack of a standardised problem definition, readily
	available benchmarking datasets, and suitable metrics. We elaborate on
	these causal reasons and provide viable solutions for 
	performing benchmarks more easily.\\
	\textbf{Conclusion:} With this perspective, we hope to address the
	elephant in the room, and provide 
	developers of trajectory inference methods with the necessary 
	tools for performing benchmarking software quantitatively. 
\end{quote}

\vfill
{\large \textbf{Publication status}}
\begin{quote}
	Manuscript in preparation. \\
	\textbf{Cannoodt R.}
\end{quote}

\vfill
{\large \textbf{Author contributions}}
\begin{quote}
	$\bullet$ \textbf{R.C.} designed the study.\\
	$\bullet$ \textbf{R.C.} surveyed 75 manuscripts from literature and analysed the data.\\
	$\bullet$ \textbf{R.C.} wrote the original manuscript.
\end{quote}

\ifrenderall
\newpage
\graphicspath{{./ch8_selfassessment/}}
\include{ch8_selfassessment/content}
\clearpage
\section{References}
\printbibliography[heading=none]
\fi

>>>>

\PassOptionsToPackage{table}{xcolor}
\documentclass[10pt]{book}

@@

%\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
%\usepackage[
%	% set width and height to a4 width and height + 6mm
%	%width=16truecm, height=24truecm,
%	a4,
%	% use any combination of these options to add different cut markings
%	cam, axes, frame, cross,
%	% set the type of TeX renderer you use
%	pdftex,
%	% center the contents
%	center
%]{crop}

\usepackage[paperheight=24cm,paperwidth=17cm,margin=1.5cm,inner=2cm,outer=1.5cm]{geometry}
% adjust margins (should be done before importing fancyhdr!) %lmargin=2.5cm,rmargin=4cm,tmargin=3cm,bmargin=3cm

@@

% sections in allcaps
\usepackage{titlesec}

@@

% magic for reading parts of a file
\makeatletter
\newread\pin@file
\newcounter{pinlineno}
\newcommand\pin@accu{}
\newcommand\pin@ext{pintmp}
% inputs #3, selecting only lines #1 to #2 (inclusive)
\newcommand*\partialinput [3] {%
	\IfFileExists{#3}{%
		\openin\pin@file #3
		% skip lines 1 to #1 (exclusive)
		\setcounter{pinlineno}{1}
		\@whilenum\value{pinlineno}<#1 \do{%
			\read\pin@file to\pin@line
			\stepcounter{pinlineno}%
		}
		% prepare reading lines #1 to #2 inclusive
		\addtocounter{pinlineno}{-1}
		\let\pin@accu\empty
		\begingroup
		\endlinechar\newlinechar
		\@whilenum\value{pinlineno}<#2 \do{%
			% use safe catcodes provided by e-TeX's \readline
			\readline\pin@file to\pin@line
			\edef\pin@accu{\pin@accu\pin@line}%
			\stepcounter{pinlineno}%
		}
		\closein\pin@file
		\expandafter\endgroup
		\scantokens\expandafter{\pin@accu}%
	}{%
		\errmessage{File `#3' doesn't exist!}%
	}%
}
\makeatother

\newcounter{acki}
\setcounter{acki}{1}
\newcommand{\doack}{%
%	{\footnotesize %
%    %\#\theacki\ | % LINENUMBER DEBUG
%		\partialinput{\theacki}{\theacki}{ack}
%	}%
	\stepcounter{acki}%
}

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE HEADER %%%
%%%%%%%%%%%%%%%%%%%%%
\usepackage{fancyhdr}  
\pagestyle{fancy} 
\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
\fancypagestyle{frontmatter}{%    
	\fancyhf{} 	
	\fancyfoot[RO,LE]{\textsf{Page \thepage}} 
	\fancypagestyle{plain}{
		\renewcommand{\headrulewidth}{0pt}
		\fancyhead{}
		\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	} 
}

\fancypagestyle{mainmatter}{%    
	\fancyhf{} 	
	\fancyhead[RO,LE]{\nouppercase{\small \textsf{\leftmark}}}
	\fancyfoot[RO,LE]{\textsf{{\footnotesize Page \thepage}}}
	\fancyfoot[LO]{\ifodd\theacki\doack\fi}
	\fancyfoot[RE]{\ifodd\theacki\else\doack\fi}

}
\setlength{\headheight}{15pt}

@@

\definecolor{light-gray}{gray}{0.90}
\definecolor{med-gray}{gray}{0.60}
\tikzset{
	thumb/.style={
		%   draw=black,
		fill=light-gray,
		text=black,
		minimum height=\thumbheight, %\thumbheight,
		text width=\thumbwidth,
		outer sep=0pt,%   outer sep=10pt,
		font=\sffamily\Huge,
	}
}
\newcommand{\oddthumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north east,] at ($%
	(current page.north east)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
\newcommand{\eventhumb}[1]{%
	\begin{tikzpicture}[remember picture, overlay]
	\node [thumb,text centered,anchor=north west,] at ($%
	(current page.north west)-%
	(0,\thumbtopmargin+\value{letternum}*\thumbheight)%
	$) {#1};
	\end{tikzpicture}
}
% create a new command to set a new lettergroup
\newcommand{\lettergroup}[1]{%
	\fancyhead[LO]{\oddthumb{#1}}%
	\fancyhead[RE]{\eventhumb{#1}}%
	\fancypagestyle{chapterstart}{%
		\renewcommand{\headrulewidth}{0pt}
		\renewcommand{\footrulewidth}{0pt}
		\fancyhf{}
		\chead{\oddthumb{#1}}% chapters start only on odd pages
		\fancyfoot[RO,LE]{\textsf{Page \thepage}}
	}
	\thispagestyle{chapterstart}
	\stepcounter{letternum}%
}

@@

\begin{document}
\pagenumbering{alph}
\pagestyle{frontmatter}
\frontmatter

@@

\newpage
\graphicspath{{./ch1_introduction/}}
\include{ch1_introduction/content}

@@

\vfill
{\large \textbf{Author contributions}}
\begin{quote}
	$\bullet$ \textbf{R.C.} and W.S. designed the study.\\
	$\bullet$ \textbf{R.C.}, W.S., and L.D. performed the experiments and analysed the data.\\
	$\bullet$ \textbf{R.C.} and W.S. implemented software packages.\\
	$\bullet$ \textbf{R.C.} and W.S. wrote the original manuscript.\\
	$\bullet$ \textbf{R.C.}, W.S., L.D., and Y.S. reviewed and edited the manuscript. \\
	$\bullet$ Y.S. supervised the project.
\end{quote}

@@

### 2019-12-10 ch0_frontmatter.tex
updates

\newpage
\ 
\vfill
{\bfseries\LARGE Summary}
\vfill
Recent developments in the \textit{single-cell omics} technologies triggered a major milestone in the field of (cell)biology. Current techniques can be used to profile both the genome and the transcriptome of individual cells, leading to important insights into cellular heterogeneity and dynamic cell processes. This evolution has led, among other things, to the Human Cell Atlas initiative, an international collaboration project with the aim of mapping all human cell types in terms of its biomolecular contents.

@@

Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects.

\newpage
%\ 
%\vfill
{\bfseries\LARGE Samenvatting}
%\vfill

Recente ontwikkelingen in de \textit{single-cell omics} technologieën vormen een belangrijke mijlpaal in het domein van de (cel)biologie. Met de huidige technieken kunnen zowel het genoom als het transcriptoom van individuele cellen geprofileerd worden wat leidt tot belangrijke inzichten in de cellulaire heterogeniteit en de dynamische celprocessen. Deze evolutie heeft onder meer geleid tot het Human Cell Atlas initiatief, een internationaal samenwerkingsproject met als doel het in kaart brengen van alle humane cell types.

@@

Gedurende dit project heb ik vooral genoten van het werken in een \textit{open-source} omgeving, en het samenwerken met nationale en internationale collega’s uit het veld om onopgeloste problemen aan te pakken.  Door de broncode van excellente softwareprojecten te bestuderen, leerde ik mezelf de aanbevolen werkwijzen aan.

>>>>

\newpage
\ 
\vfill
{\bfseries\LARGE Summary}
\vfill
{
\fontsize{9}{10}\selectfont
Recent developments in the \textit{single-cell omics} technologies triggered a major milestone in the field of (cell)biology. Current techniques can be used to profile both the genome and the transcriptome of individual cells, leading to important insights into cellular heterogeneity and dynamic cell processes. This evolution has led, among other things, to the Human Cell Atlas initiative, an international collaboration project with the aim of mapping all human cell types in terms of its biomolecular contents.

@@

Throughout this project I particularly enjoyed working in an open-source ecosystem, collaborating with scientists around the world to tackle unsolved problems, whilst learning best practices by inspecting the source code of exemplary software projects. 
\vfill
}

\newpage
\ 
\vfill
{\bfseries\LARGE Samenvatting}
\vfill

{
\fontsize{9}{10}\selectfont
Recente ontwikkelingen in de \textit{single-cell omics} technologieën vormen een belangrijke mijlpaal in het domein van de (cel)biologie. Met de huidige technieken kunnen zowel het genoom als het transcriptoom van individuele cellen geprofileerd worden wat leidt tot belangrijke inzichten in de cellulaire heterogeniteit en de dynamische celprocessen. Deze evolutie heeft onder meer geleid tot het Human Cell Atlas initiatief, een internationaal samenwerkingsproject met als doel het in kaart brengen van alle humane cell types.

@@

Gedurende dit project heb ik vooral genoten van het werken in een \textit{open-source} omgeving, en het samenwerken met nationale en internationale collega’s uit het veld om onopgeloste problemen aan te pakken.  Door de broncode van excellente softwareprojecten te bestuderen, leerde ik mezelf de aanbevolen werkwijzen aan.
\vfill
}

### 2019-12-10 ch3_dynbenchmark/content.tex
updates

\newpage
\subsection{Accuracy}

@@

\begin{table}[tbh!]
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}

@@

	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
		\toprule
		Aspect & Items & References \\ \midrule

@@

		\midrule
		\multicolumn{3}{l}{\textbf{Documentation}} \\*

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (\url{https://methods.dynverse.org}). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

The dyngen (\href{https://github.com/dynverse/dyngen}{https://github.com/dynverse/dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. We simulated a total of 110 datasets, with 11 different topologies.

\hfill\break
\textit{Network generation}
\hfill\break

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcation seems to be mutual antagonism, where genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}. Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}. Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.

To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters. Two module networks generated linear trajectories (linear and linear long), one generated a bifurcation, one generated a convergence, one generated a multifurcation (trifurcating), two generated a tree (consecutive bifurcating and binary tree), one generated an acyclic graph (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple graph structures (bifurcating loop and bifurcating cycle). The structure of these module networks is available at \href{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}{https://github.com/dynverse/dyngen/tree/master/inst/ext\_data/modulenetworks}.

From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5$\%$ of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits \cite{marbach_tissuespecificregulatorycircuits_2016}), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the page\_rank function of the \textit{igraph} package.

\hfill\break
\textit{Simulation of gene regulatory systems using thermodynamic models}
\hfill\break

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods \cite{marbach_wisdomcrowdsrobust_2012}. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations \cite{schaffter_genenetweaversilicobenchmark_2011} (ODEs):

$$
\begin{aligned}
\label{eq:mrna_ode}
\frac{dx_i}{dt} &= \underbrace{m \times f(y_1, y_2, ...)}_\text{production} - \underbrace{\lambda \times x_i}_\text{degradation}
\end{aligned}
$$
$$
\begin{aligned}
\label{eq:prot_ode}
\frac{dy_i}{dt} &= \underbrace{r \times x_i}_\text{production} - \underbrace{\Lambda \times y_i}_\text{degradation}
\end{aligned}
$$

where $m$, $\lambda$, $r$ and $\Lambda$ represent production and degradation rates, the ratio of which determines the maximal gene and protein expression. The two types of equations are coupled because the production of protein $y_i$ depends on the amount of gene expression $x_i$, which in turn depends on the amount of other proteins through the activation function $f(y_1, y_2, ...)$.

The activation function is inspired by a thermodynamic model of gene regulation, in which the promoter of a gene can be bound or unbound by a set of transcription factors, each representing a certain state of the promoter. Each state is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state. The production rate of the gene is calculated by combining the probabilities of the promoter being in each state with the relative activation:

$$
\begin{aligned}
\label{eq:input_function_probabilities}
f(y_1, y_2, ..., y_n) = \sum_{j \in \{0, 1, ..., n^2\}} \alpha_j \times P_j
\end{aligned}
$$

The probability of being in a state is based on the thermodynamics of transcription factor binding. When only one transcription factor is bound in a state:
$$
\begin{aligned}
P_j \propto \nu = \left(\frac{y}{k}\right)^{n}
\end{aligned}
$$

where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
$$
\begin{aligned}
P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
\end{aligned}
$$

where $\rho$ represents the cooperativity of binding between the different transcription factors.

$P_i$ is only proportional to $\nu$ because $\nu$ is normalized such that $\sum_{i} P_i = 1$.

To each differential equation, we added an additional stochastic term: 
$$
\begin{aligned}
\frac{dx_i}{dt} & = m \times f(y_1, y_2, ...) - \lambda \times x_i + \eta \times \sqrt{x_i} \times \Delta W_t \\
\frac{dy_i}{dt} & = r \times x_i - \Lambda \times y_i + \eta \times \sqrt{y_i} \times \Delta W_t
\end{aligned}
$$

with $\Delta W_t \sim \mathcal{N}(0, h)$.

Similar to GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011}, we sample the different parameters from random distributions, defined as follows. $e$ defines whether a transcription factor activates (1) or represses (-1), as defined within the regulatory network network.

$$
\begin{aligned}
r & = \mathcal{U}(10, 200) \\
d & = \mathcal{U}(2, 8) \\
p & = \mathcal{U}(2, 8) \\
q & = \mathcal{U}(1, 5) \\
a_0 & = \begin{cases}1 & \text{if } |e| = 0 \\ 1 & \text{if } \forall x \in e \text{, } x = -1 \\ 0 & \text{if } \forall x \in e \text{, } x = 1 \\ 0.5 & \text{otherwise}\end{cases} \\
a_i & = \begin{cases}0 & \text{if } \exists x \in e_i \text{, } x = -1 \\ 1 & \text{otherwise}\end{cases} \\
s & = \mathcal{U}(1, 20) \\
k & = y_{max} / (2 * s) \text{,} \\
\ & \quad \text{where } y_{max} = r / d \times p / q \\
c & = \mathcal{U}(1, 4)
\end{aligned}
$$

We converted each ODE to an SDE by adding a chemical Langevin equation, as described in \cite{schaffter_genenetweaversilicobenchmark_2011}. These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$. The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets. The burn-in period was for each simulation 2. Each network was simulated 32 times.

\hfill\break
\textit{Simulation of the single-cell RNA-seq experiment}
\hfill\break

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. These cells were sampled uniformly across the different steps of the 32 simulations.  Next, we used the Splatter package \cite{zappia_splattersimulationsinglecell_2017} to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$. Both $x_0$ (the midpoint for the dropout logistic function) and $k$ (the shape of the dropout logistic function) were estimated by Splatter.

\hfill\break
\textit{Gold standard extraction}
\hfill\break

Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable. We therefore first constructed a backbone expression profile for each branch within the trajectory. To do this, we first defined in which order the expression of the modules is expected to change, and then generated a backbone expression profile in which the expression of these modules increases and decreases smoothly between 0 and 1.  We also smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation.  We used dynamic time warping, implemented in the dtw R package \cite{giorgino_computingvisualizingdynamic_2009,tormene_matchingincompletetime_2009}, with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

@@

\begin{itemize}
	\item A user may be primarily interested in defining the correct topology, and only use the cellular ordering when the topology is correct
	\item A user may be less interested in how the cells are ordered within a branch, but primarily in which cells are in which branches
	\item A user may already know the topology, and may be primarily interested in finding good features related to a particular branching point
	\item ...
\end{itemize}

>>>>

\subsection{Accuracy}

@@

\begin{table}[tbh!]
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development. Part 1.} \label{tab:scoresheet}

@@

	\begin{tabular}{p{2.5cm}p{8cm}p{1.5cm}}
		\toprule
		Aspect & Items & References \\ \midrule

@@

\begin{table}[tbh!]
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Part 2.} \label{tab:scoresheet2}

	\begin{tabular}{p{2.5cm}p{8cm}p{1.5cm}}
		\toprule
		Aspect & Items & References \\ \midrule

		\multicolumn{3}{l}{\textbf{Documentation}} \\*

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (\url{https://methods.dynverse.org}). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{lamanno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

The dyngen (Chapter~\ref{chap:dyngen}) workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012} and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we tried to mirror real regulatory networks, while keeping the model simple and easily extendable. We simulated a total of 110 datasets, with 11 different topologies.

@@

\begin{itemize}
	\item A user may be primarily interested in defining the correct topology, and only use the cellular ordering when the topology is correct
	\item A user may be less interested in how the cells are ordered within a branch, but primarily in which cells are in which branches
	\item A user may already know the topology, and may be primarily interested in finding good features related to a particular branching point
\end{itemize}
